\documentclass{article}
\usepackage{graphics}
\usepackage{booktabs}
\usepackage{colortbl, xcolor}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{url}
%\usepackage{natbib}
\usepackage{epstopdf}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage[a4paper, total={6in, 8in}]{geometry}

\author{Micha\l{} Burdukiewicz, Przemys\l{}aw Gagat, Micha\l{} Gaworski, \\
S\l{}awomir Jab\l{}o\'nski, Pawe\l{} Mackiewicz, Marcin \L{}ukaszewicz}
\date{}
\title{PhyMet$^2$: phylogenetic tree \\
\large Supplemental materials}
\renewcommand{\section}[2]{}%

\begin{document}

\maketitle
    The 16S rRNA sequences of methanogens were extracted from GenBank 
(\url{https://www.ncbi.nlm.nih.gov/}) and Silva (\url{https://www.arb-silva.de/}) databases. 
The sequences were aligned in MAFFT using slow and accurate algorithm L-INS-i 
with 1,000 cycles of iterative refinement. Sites suitable for phylogenetic 
studies from the alignments were selected in GBlocks . The phylogenetic tree was 
constructed using MrBayes . We applied the mixed model of nucleotide 
substitutions to specify appropriate substitution models across the large 
parameter space . Moreover, we assumed the heterogeneity rate across sites 
described by the invariant model and the discrete gamma model with five category 
rates. Two independent runs starting from random trees, each using eight Markov 
chains, were applied. The trees were sampled every 100 generations for 
20,000,000 generations. In the final analysis, we selected the last 85,720 trees 
that reached the stationary phase and convergence (i.e. when the standard 
deviation of split frequencies stabilized and was lower than the proposed 
threshold of 0.01).


\begin{thebibliography}{9}
\bibitem{b1} 
Katoh K, Standley DM:
\textit{MAFFT multiple sequence alignment software version 7: 
improvements in performance and usability}. 
Molecular biology and evolution 2013, 30(4):772-780.
 
\bibitem{b2} 
Talavera G, Castresana J:
\textit{Improvement of phylogenies after removing 
divergent and ambiguously aligned blocks from protein sequence alignments}. 
Systematic biology 2007, 56(4):564-577.

\bibitem{b3} 
Ronquist F, Teslenko M, van der Mark P, Ayres DL, Darling A, Hohna S, 
Larget B, Liu L, Suchard MA, Huelsenbeck JP:
\textit{MrBayes 3.2: efficient Bayesian 
phylogenetic inference and model choice across a large model space}. 
Systematic biology 2012, 61(3):539-542.

\bibitem{b4} 
Huelsenbeck JP, Larget B, Alfaro ME:
\textit{Bayesian phylogenetic model 
selection using reversible jump Markov chain Monte Carlo}. Molecular biology and 
evolution 2004, 21(6):1123-1133

\end{thebibliography}


\end{document}